Predicting the spread and persistence of genetically modified dominant sterile male mosquitoes

### 2021-02-05 inst/manuscript/Draft/bmc_article_v10.tex
2nd round of tm-fnih comments

% latex table generated in R 3.4.1 by xtable 1.8-2 package
% Fri Oct 13 08:25:21 2017
\begin{table}[ht]
\centering
\begingroup\small
\begin{tabular}{|cccccc|}
\hline
MRR & Date & Release site & Longitude & Latitude & Number released \\ 
\hline
1 & 2013-10-09 & A & -4.4724 & 11.2347 & 1146 \\ 
1 & - & B & -4.4755 & 11.2342 & 1158 \\ 
1 & - & C & -4.4718 & 11.2318 & 1103 \\ 
2 & 2014-05-07 & A & -4.4724 & 11.2347 & 1878 \\ 
2 & - & B & -4.4755 & 11.2342 & 1655 \\ 
2 & - & C & -4.4718 & 11.2318 & 1734 \\ 
3 & 2014-09-04 & A & -4.4724 & 11.2347 & 1665 \\ 
3 & - & B & -4.4755 & 11.2342 & 1673 \\ 
3 & - & C & -4.4718 & 11.2318 & 1684 \\ 
4 & 2015-04-09 & A & -4.4724 & 11.2347 & 2107 \\ 
4 & - & B & -4.4755 & 11.2342 & 2013 \\ 
4 & - & C & -4.4718 & 11.2318 & 1953 \\ 
5 & 2015-10-09 & C & -4.4724 & 11.2347 & 5992 \\ 
\hline
\end{tabular}
\endgroup
\caption{Summary of the MRR experiments conducted by \cite{Epopa2017} in Burkina Faso. We use the results of the first four experiments to parameterise our model and calculate Bayes factor for expert-derived priors, and use the fifth MRR experiment is to validate the model predictions.} 
\label{tab:relLoc}
\end{table}

@@

\begin{table}[h]
\centering
\begingroup\small
\begin{tabular}{|c|ccc|}
\hline
Parameter & Prior parameters & Prior distribution & Source \\\hline
Mortality ($\mu$) & $ \theta_\mu = \{\alpha_{\mu}, \beta_{\mu}\}$ & Beta & Experts\\
Diffusion ($D$) & $\theta_D = \{ m_D, \sigma_D^2$ \} & Log-normal & Experts\\
Swarm attraction ($\alpha$) & $\theta_\alpha = \{ m_\alpha, \sigma_\alpha^2 \}$ & Log-normal &  Weakly informative \\ 
Swarm range ($\sigma$) &  $\theta_\sigma = \{m_\sigma, \sigma_\sigma^2 \}$ & Log-normal &  Literature\\ 
Catchability cfr ($p_{cfr}$) & $\theta_{cfr} =\{\alpha_{cfr}, \beta_{cfr}\}$ & Beta & Weakly informative \\ 
Catchability psc ($p_{psc}$) & $\theta_{psc} =\{\alpha_{psc}, \beta_{psc}\}$ & Beta & SOP \\ 
Catchability ss ($p_{ss}$) & $\theta_{ss} =\{\alpha_{ss}, \beta_{ss}\}$ & Beta & SOP \\
\hline
\end{tabular}
\endgroup
\caption{\label{tble:prior} Process (PDE model) and observation model parameters, prior distributions and sources. SOP: Standard Operating Procedures: the procedure used during the MRR experiments are described in a set of SOP that were made available to us as part of an independent risk assessment process.}
\end{table}

@@

\section{Results}
\subsection{WT parameters}
Table \ref{tble:sumStatPPpost} provides the summary statistics of the posterior distributions of the PDE model parameters (Figure \ref{fig:prior_posterior} provides both the priors and posteriors for comparison). The use of the MRR data allowed us to provide refined estimates for the parameters of the PDE model. The diffusion coefficient in particular has seen its uncertainty decrease, yielding a mean value of 127 m$^2$ per day. The chemotaxis component has a posterior mean value of about $0.07$, while the range parameter for the mosquito attraction is about $33.9$ m.\\

@@

\begin{table}[h]
\centering
\begingroup\small
\begin{tabular}{|c|ccc|}
\hline
Parameter & Mean & Q05 & Q95 \\\hline
Mortality ($\mu$) & 0.16 & 0.11 & 0.24\\
Diffusion ($D$) & 127.0 & 113.7 & 140.7\\
Swarm attraction ($\alpha$) & 0.07 & 0.03 & 0.10\\
Swarm range ($\sigma$) & 33.9 & 19.8 & 56.1\\
Catchability psc ($p_{psc}$) & 0.18 & 0.11 & 0.25\\
Catchability cfr ($p_{cfr}$) & 0.03 & 0.02 & 0.05\\
Catchability ss ($p_{ss}$) & 0.29 & 0.24 & 0.34\\
\hline
\end{tabular}
\endgroup
\caption{\label{tble:sumStatPPpost} Summary statistics for the posterior distributions of the PDE model parameters inferred from the WT MRRs. }
\end{table}

@@

\begin{table}[h]
\centering
\begingroup\small
\begin{tabular}{|c|ccc|}
\hline
Parameter & Mean & Q05 & Q95  \\\hline
Mortality ($\mu$) & 1.05 & 0.19 & 2.42\\
Diffusion ($D$) & 753 $\times$ 10$^{3}$ & 8.77 & 112  $\times$ 10$^{4}$\\
\hline
\end{tabular}
\endgroup
\caption{\label{tble:sumStatPPpostbma} Updated statistics for the PDE model parameters using the WT MRRs results and the DSM priors. }
\end{table}

@@

The dispersal of the DSM mosquitoes is balanced by the mortality rate. Because of this, the DSM are not really expected to disperse far from their release site, with the expected number of mosquitoes per cell delineated by the red and orange contours in Figure \ref{fig:plotSpread}. For instance, while the spread is expected to extend on average to up to $500$m away from the release location by day $2$ (with $1\%$ chance that it extends further), the extent then contracts quickly to a limited area (less than $250$m from release location) by day $5$, to finally about $150$m away from the release location at day $12$.

@@

\section{Discussion}
In this analysis we have used a combination of mathematical modeling, Bayesian inference, and expert elicitation to predict the spread and persistence of genetically modified DSM mosquitoes following the release of a single cohort of $5000$ individuals. These predictions were made as part of an independent risk assessment that was completed in May 2018 prior to the proposed field release \citep{Hayes2015a, Hayes2018}. The field release itself occurred in July 2019 (\url{https://targetmalaria.org/target-malaria-proceeded-with-a-small-scale-release-of-genetically-modified-sterile-male-mosquitoes-in-bana-a-village-in-burkina-faso/}). We anticipate that the results from this field experiment will soon be forthcoming and provide an important opportunity to test the accuracy of the predictions presented here.

>>>>

% latex table generated in R 3.4.1 by xtable 1.8-2 package
% Fri Oct 13 08:25:21 2017
\begin{table}[ht]
\centering
\begingroup\small
\begin{tabular}{|cccccccc|}
\hline
MRR & Date & Release site & Longitude & Latitude & N released & N captured & Distance\\ 
\hline
1 & 2013-10-09 & A & -4.4724 & 11.2347 & 1146 & 32 & 140\\ 
1 & - & B & -4.4755 & 11.2342 & 1158 & 9 & 344\\ 
1 & - & C & -4.4718 & 11.2318 & 1103 & 6 & 205\\ 
2 & 2014-05-07 & A & -4.4724 & 11.2347 & 1878 & 6 & 94\\ 
2 & - & B & -4.4755 & 11.2342 & 1655 & 1 & 266\\ 
2 & - & C & -4.4718 & 11.2318 & 1734 & 1 & 193\\ 
3 & 2014-09-04 & A & -4.4724 & 11.2347 & 1665 & 56 & 133\\ 
3 & - & B & -4.4755 & 11.2342 & 1673 & 4 & 440\\ 
3 & - & C & -4.4718 & 11.2318 & 1684 & 13 & 205\\ 
4 & 2015-04-09 & A & -4.4724 & 11.2347 & 2107 & 1 & 55\\ 
4 & - & B & -4.4755 & 11.2342 & 2013 & 3 & 386\\ 
4 & - & C & -4.4718 & 11.2318 & 1953 & 18 & 190\\ 
5 & 2015-10-09 & C & -4.4724 & 11.2347 & 5992 & 18 & 141\\ 
\hline
\end{tabular}
\endgroup
\caption{Summary of the MRR experiments conducted by \cite{Epopa2017} in Burkina Faso. The last column presents the average distance for recapture (in meters). We use the results of the first four experiments to parameterise our model and calculate Bayes factor for expert-derived priors, and use the fifth MRR experiment is to validate the model predictions.} 
\label{tab:relLoc}
\end{table}

@@

\begin{table}[h]
\centering
\begingroup\small
\begin{tabular}{|c|ccc|}
\hline
Parameter & Prior parameters & Prior distribution & Source \\\hline
Mortality ($\mu$) & $ \theta_\mu$: mixture (see Fig \ref{fig:prior_posterior}) & Beta & Experts\\
Diffusion ($D$) & $\theta_D$: mixture (see Fig \ref{fig:prior_posterior}) & Log-normal & Experts\\
Swarm attraction ($\alpha$) & $\theta_\alpha = \{ -2, 1 \}$ & Log-normal &  Weakly informative \\ 
Swarm range ($\sigma$) &  $\theta_\sigma = \{3.5, 4 \}$ & Log-normal &  Literature\\ 
Catchability cfr ($p_{cfr}$) & $\theta_{cfr} =\{1, 1\}$ & Beta & Weakly informative \\ 
Catchability psc ($p_{psc}$) & $\theta_{psc} =\{1.4, 1\}$ & Beta & SOP \\ 
Catchability ss ($p_{ss}$) & $\theta_{ss} =\{10, 20\}$ & Beta & SOP \\
\hline
\end{tabular}
\endgroup
\caption{\label{tble:prior} Process (PDE model) and observation model parameters, prior distributions and sources. SOP: Standard Operating Procedures: the procedure used during the MRR experiments are described in a set of SOP that were made available to us as part of an independent risk assessment process.}
\end{table}

@@

\section{Results}
\subsection{WT parameters}
Table \ref{tble:sumStatPPpost} provides the summary statistics of the posterior distributions of the PDE model parameters (Figure \ref{fig:prior_posterior} provides both the priors and posteriors for comparison). The use of the MRR data allowed us to provide refined estimates for the parameters of the PDE model. The diffusion coefficient in particular has seen its uncertainty decrease, yielding a mean value of 127 m$^2$ per day. The chemotaxis component has a posterior mean value of about $0.07$, while the range parameter for the mosquito attraction is about $33.9$ m.\\
In the posterior samples, we could identify a high level of correlation between the dispersal and swarm attraction parameters, as can be expected. The shape of this correlation means that high dispersal values correlates with low attraction, and vice-versa. The posterior distribution of dispersal and attraction are also linked to the range of the traps laid out in the field. With limited long range traps, no long range catches were made, limiting the ability of the model to capture possible high dispersal values.

@@

\begin{table}[h]
\centering
\begingroup\small
\begin{tabular}{|c|ccc|}
\hline
Parameter & Mean & Q05 & Q95 \\\hline
Mortality - $\mu$ & 0.16 (0.14) & 0.11 (0.10) & 0.24 (0.21)\\
Diffusion - $D$ & 127.0 & 113.7 & 140.7\\
Swarm attraction - $\alpha$ & 0.07 & 0.03 & 0.10\\
Swarm range - $\sigma$ & 33.9 & 19.8 & 56.1\\
Catchability psc - $p_{psc}$ & 0.18 & 0.11 & 0.25\\
Catchability cfr - $p_{cfr}$ & 0.03 & 0.02 & 0.05\\
Catchability ss - $p_{ss}$ & 0.29 & 0.24 & 0.34\\
\hline
\end{tabular}
\endgroup
\caption{\label{tble:sumStatPPpost} Summary statistics for the posterior distributions of the PDE model parameters inferred from the WT MRRs. In brackets in the first line we added the equivalent daily mortality rate value, for ease of comparison with other studies.}
\end{table}

@@

\begin{table}[h]
\centering
\begingroup\small
\begin{tabular}{|c|ccc|}
\hline
Parameter & Mean & Q05 & Q95  \\\hline
Mortality - $\mu$ & 1.05 (0.65) & 0.19 (0.17) & 2.42 (0.91)\\
Diffusion - $D$ & 753 $\times$ 10$^{3}$ & 8.77 & 112  $\times$ 10$^{4}$\\
\hline
\end{tabular}
\endgroup
\caption{\label{tble:sumStatPPpostbma} Updated statistics for the PDE model parameters using the WT MRRs results and the DSM priors. In brackets in the first line we added the equivalent daily mortality rate value, for ease of comparison with other studies.}
\end{table}

@@

The dispersal of the DSM mosquitoes is balanced by the mortality rate. Because of this, the DSM are not really expected to disperse far from their release site, with the expected number of mosquitoes per cell delineated by the red and orange contours in Figure \ref{fig:plotSpread}. For instance, while the spread is expected to extend on average to up to $500$m away from the release location by day $2$ (there is a probability of $0.99$ or more  of not finding a single mosquito further away), the extent then contracts quickly to a limited area (less than $250$m from release location) by day $5$, to finally about $150$m away from the release location at day $12$.

@@

\section{Discussion}
In this analysis we have used a combination of mathematical modeling, Bayesian inference, and expert elicitation to predict the spread and persistence of genetically modified DSM mosquitoes following the release of a single cohort of $5000$ individuals. These predictions were made as part of an independent risk assessment that was completed in May 2018 prior to the proposed field release \citep{Hayes2015a, Hayes2018}. The field release itself occurred in July 2019 (\href{https://tinyurl.com/y4qr9lc5}{See here}). We anticipate that the results from this field experiment will soon be forthcoming and provide an important opportunity to test the accuracy of the predictions presented here.
